Palomar 5 Gaps

### 2024-10-21 paper/main.tex
more citations and little changes

% \usepackage[draft]{graphicx}
% \usepackage{txfonts}
\usepackage{amsmath}

@@

\newcommand{\paola}[1]{\textcolor{magenta}{{#1}}}
\newcommand{\salvatore}[1]{\textcolor{green}{{#1}}}
\newcommand{\marcook}[1]{\textcolor{green}{{#1}}}
\newcommand{\marcoko}[1]{\textcolor{red}{{#1}}}
\newcommand{\qmar}[1]{\textcolor{brown}{{#1}}}

@@

  \salvatore{I couldn't find the bib tex for zwicky's original paper}
  Over a century ago, \citet{1933AcHPh...6..110Z} postulated the existence of dark matter to explain the excessively large velocity dispersions of galaxies within their galaxy clusters. Since then, mounting evidence has been assembled to support this claim. \citet{1970IAUS...38..351F,1970ApJ...161..802F} elaborated on the dynamics of disk galaxies and briefly acknoledged in the appendix of his paper that the rotation curves of NGC 300 and M~33 cannot be fully explained by disk models and that there must be the same amount of mass or more that is undetected and beyond the extend of the galaxies to explain their rotation curves. Concurrently \citet{1970ApJ...159..379R} studied the rotation of Andromenda and observed it to be flat and not decreasing as one would expect from a potential with more mass concentration in the center. Eventually, astronomers returned to Zwicky's non-luminous matter to explain these fast-rotating disks, as discussed in \citet{1983SciAm.248f..96R}'s review.

  \salvatore{I see why we cite freeman70 now. there are two comments in the appendix stating that there is as much mass beyond the optical limit as there is within it for two different galaxoes. NGC 300 and M 33. It's not clear to me why we cite Rubin 70. The rotation curve is flat, to me implies a logarithmic potential and hence we need dark matter. However, I don't get any impression of missing matter when I read the text.}

  Beyond the galactic scale, dark matter is essential for explaining many cosmological phenomena. For instance, the Planck mission \citep{2020A&A...641A...1P} was designed to study the Cosmic Microwave Background (CMB), whose spectrum closely follows that of a black body with a peak temperature in the millimeter wavelength range, corresponding to approximately 2.725~K. However, there are slight temperature variations on the order of one part in ten thousand. Notably, the colder regions of the CMB temperature map correlate with whole-sky gravitational lensing maps \citep{2020A&A...641A...8P}. This is because photons lose energy as they escape from gravitational potentials. Additionally, galaxy clusters reside within these deep gravitational wells. The $\Lambda$ Cold Dark Matter model accounts for these observations almost perfectly, proposing that dark matter exists and that its stochastic quantum fluctuations led to density variations. These variations propagated as the universe expanded, shaping the large-scale structure and distribution of galaxies we observe today.

  As laid out in \citet{2021PrPNP.11903865A}'s review, despite significant progress, the true nature of dark matter remains elusive and scientists continue to search for evidence of its existence beyond its gravitational effects on cosmic and galactic scales. A key focus today is detecting dark matter on sub-galactic scales, because at these scales different cosmological scenarios predict different lumpiness of the halos. For instance, \citet{2008ApJ...679.1260M} compared the Via Lactea simulations to some that they made themselves and showed that you can have a large variation for the number of dark matter subhalos within a galaxy. \citet{2019Galax...7...81Z} provides a review for the different proposed mechanisms of forming dark matter haloes from perturbation theory and deviations from a uniformly dense medium and also disccses techniques and limitations for modelign such physical phenomena. \citet{2019Galax...7...81Z} rightly states that until particles are discovered $\Lambda$-CDM remains a hypothesis. Since the true nature of dark matter is elusive. The possibilities for sub-galactic behaviour and structure of dark matter are vaste. However, as proposed by Carlberg in 2012, a most promising tool to set limits on the number and mass density of the sub-halos, are stellar streams.

@@

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in Ferrone et al. (2023), which we present here for completeness. We begin by extracting the phase space properties (positions, velocities) as well as the masses and half-mass radii of 165 globular clusters from the galactic globular cluster catalog by Baumgardt and Vasiliev et al. We handle the uncertainties on the measured phase space parameters through a monte-carlo approach. The catalog provides uncertainties for mass, distances, line of sight velocities, proper motions and includes a covariance term between the proper motions---uncertainties for the right ascension, declination, and the characterisitcs radii are not available. We repeate simulation the creation of Palomar 5's tidail tails 50 times, for each of which we vary the initial conditions of each cluster by sampling its five-dimensional gaussian uncertaintiy distribution. For the first simulation, which we name \texttt{Sampling 000}, we use the most probable values for the initial conditions.

    We convert the initial conditions from sky coordinates into a galactocentric reference frame. We use the local standard of rest and the peculiary motion of the sun as reported by Sch\"onrich 2010. The circular velocity at the solar radius was set to: $v_{\text{LSR}} = 240$~km~s$^{-1}$. The peculiar velocity of the sun was set to $(U_\odot, V_\odot, W_\odot)=(11.1, 12.24, 7.25)$~km~s$^{-1}$. The position of the Sun was set to $(x_\odot,y_\odot,z_\odot) = (-8.34,0,0.027)$~kpc, vertical position above the disk was taken from Chen et al 2001 and the galacto-radial distance was taken from Reid et al 2014. These transformations were performed using the \texttt{astropy} \citep{2013A&A...558A..33A}.

@@

    At their computed positions 5 Gyr ago, each globular cluster was assigned 100,000 particles sampled from a Plummer distribution to represent its stellar content. For each set of initial conditions, the particle positions were resampled to account for uncertainties in mass, which is an improvement from \citet{2023A&A...673A..44F}. Since globular clusters are older than 8 Gyr, their most massive stars have already gone supernova, leaving primarily low-mass stars, typically under one solar mass (CITATION?). Given that clusters usually have masses between $10^5-10^6$ solar masses, 100~000 particles is sufficient to represent the distribution of tidal debris.

@@

    The integration was performed using a leapfrog integrator (add citation). The timestep size was chosen to be based on two critera. First, to ensure the conservation of energy to XXX. The second, is the drift in position due to numerical error over 5 Gyr was negligible compared to the scale radius of the globular clusters, which is about 5~pc. Further details on the timestep selection are discussed in the appendix.

@@

    Overall, across 50 simulations, we observe 73 gaps. The distribution of simulations with a given number of gaps is as follows: 0 gaps in 4 simulations, 1 gap in 25 simulations, 2 gaps in 16 simulations, 3 gaps in 4 simulations, and 4 gaps in 1 simulation. Excluding NGC~2808, we are left with 29 gaps, leading to an expected range of 0.6-1.5 gaps per simulation.

>>>>

% \usepackage[draft]{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{xcolor}

@@

\definecolor{darkgreen}{rgb}{0.0, 0.5, 0.0}

\newcommand{\paola}[1]{\textcolor{magenta}{{#1}}}
\newcommand{\salvatore}[1]{\textcolor{darkgreen}{{#1}}}
\newcommand{\marcook}[1]{\textcolor{green}{{#1}}}
\newcommand{\marcoko}[1]{\textcolor{red}{{#1}}}
\newcommand{\qmar}[1]{\textcolor{brown}{{#1}}}

@@

  Over a century ago, \citet{1933AcHPh...6..110Z} postulated the existence of dark matter to explain the excessively large velocity dispersions of galaxies within their galaxy clusters. Since then, mounting evidence has been assembled to support this claim. \citet{1970IAUS...38..351F,1970ApJ...161..802F} elaborated on the dynamics of disk galaxies and briefly acknoledged in the appendix of his paper that the rotation curves of NGC 300 and M~33 cannot be fully explained by disk models and that there must be the same amount of mass or more that is undetected and beyond the extend of the galaxies to explain their rotation curves. Concurrently \citet{1970ApJ...159..379R} studied the rotation of Andromenda and observed it to be flat and not decreasing as one would expect from a potential with more mass concentration in the center. Eventually, astronomers returned to Zwicky's non-luminous matter to explain these fast-rotating disks, as discussed in \citet{1983SciAm.248f..96R}'s review. \salvatore{I see why we cite freeman70 now. there are two comments in the appendix stating that there is as much mass beyond the optical limit as there is within it for two different galaxoes. NGC 300 and M 33. It's not clear to me why we cite Rubin 70. The rotation curve is flat, to me implies a logarithmic potential and hence we need dark matter. However, I don't get any impression of missing matter when I read the text.}

  Beyond the galactic scale, dark matter is essential for explaining many cosmological phenomena. For instance, the Planck mission \citep{2020A&A...641A...1P} was designed to study the Cosmic Microwave Background (CMB), whose spectrum closely follows that of a black body with a peak temperature in the millimeter wavelength range, corresponding to approximately 2.725~K. However, there are slight temperature variations on the order of one part in ten thousand. Notably, the colder regions of the CMB temperature map correlate with whole-sky gravitational lensing maps \citep{2020A&A...641A...8P}--since the photons lost energy as they escaped the gravitational wells. The $\Lambda$ Cold Dark Matter ($\Lambda$-CDM) model accounts for these observations almost perfectly, proposing that dark matter exists and that its distribution originated from stochastic quantum fluctuations within the first second of the universe led to density variations. These variations propagated as the universe expanded, shaping the large-scale structure and distribution of galaxies we observe today.

  As laid out in \citet{2021PrPNP.11903865A}'s review, despite significant progress, the true nature of dark matter remains elusive and scientists continue to search for evidence of its existence beyond its gravitational effects on cosmic and galactic scales. A key focus today is detecting dark matter on sub-galactic scales, because at these scales different cosmological scenarios predict different lumpiness of the halos. For instance, \citet{2008ApJ...679.1260M} computed cosmological simulations and compared the Via Lactea simulations \citep{2008ApJ...679.1260M} and demonstrated (\salvatore{I don't quite understand Madau 2008. The main result is that based on the assumptions you can produce a variance in the number of subhaloes that are consistent with global properties of a milky-way like galaxy? I need to discuss this with another human}). \citet{2019Galax...7...81Z} provides a review for the different proposed mechanisms of forming dark matter haloes from perturbation theory and deviations from a uniformly dense medium and also disccses techniques and limitations for modelign such physical phenomena. \citet{2019Galax...7...81Z} rightly states that until particles are discovered $\Lambda$-CDM remains a hypothesis. Since the true nature of dark matter is elusive. The possibilities for sub-galactic behaviour and structure of dark matter are vaste. However, as proposed by Carlberg in 2012, a most promising tool to set limits on the number and mass density of the sub-halos, are stellar streams.

@@

  \subsection{Numerical Methodology}
    Our numerical methodology follows the approach outlined in Ferrone et al. (2023), which we present here for completeness. We begin by extracting the phase space properties (positions, velocities) as well as the masses and half-mass radii of 165 globular clusters from the galactic globular cluster catalog by \cite{2021MNRAS.505.5957B}.\footnote{The Baumgardt catalog has been assembled across a series of works, see: \cite{2020PASA...37...46B,2019MNRAS.482.5138B,2018MNRAS.478.1520B} as well. The catalog can be found on the World Wide Web at \href{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}{https://people.smp.uq.edu.au/HolgerBaumgardt/globular/}.} We handle the uncertainties on the measured phase space parameters through a monte-carlo approach. The catalog provides uncertainties for mass, distances, line of sight velocities, proper motions and includes a covariance term between the proper motions---uncertainties for the right ascension, declination, and the characterisitcs radii are not available. We repeate simulation the creation of Palomar 5's tidail tails 50 times, for each of which we vary the initial conditions of each cluster by sampling its five-dimensional gaussian uncertaintiy distribution. For the first simulation, which we name \texttt{Sampling 000}, we use the most probable values for the initial conditions.

    We convert the initial conditions from sky coordinates into a galactocentric reference frame. We use the local standard of rest and the peculiary motion of the sun as reported by Sch\"onrich 2010. The circular velocity at the solar radius was set to: $v_{\text{LSR}} = 240$~km~s$^{-1}$. The peculiar velocity of the sun was set to $(U_\odot, V_\odot, W_\odot)=(11.1, 12.24, 7.25)$~km~s$^{-1}$. The position of the Sun was set to $(x_\odot,y_\odot,z_\odot) = (-8.34,0,0.027)$~kpc, vertical position above the disk was taken from Chen et al 2001 and the galacto-radial distance was taken from Reid et al 2014. These transformations were performed using \texttt{astropy} \citep{2013A&A...558A..33A}.

@@

    At their computed positions 5 Gyr ago, each globular cluster was assigned 100,000 particles sampled from a Plummer distribution to represent its stellar content. For each set of initial conditions, the particle positions were resampled to account for uncertainties in mass, which is an improvement from \citet{2023A&A...673A..44F}. Since globular clusters are older than 8 Gyr, their most massive stars have already gone supernova, leaving primarily low-mass stars, typically under one solar mass (\@\salvatore{Citation? Paolo knows this stuff well.}). Given that clusters usually have masses between $10^5-10^6$ solar masses, 100~000 particles is sufficient to represent the distribution of tidal debris.

@@

    The integration was performed using a leapfrog integrator. The timestep size was chosen to be based on two critera. First, to ensure the conservation of energy to XXX. The second, is the drift in position due to numerical error over 5 Gyr was negligible compared to the scale radius of the globular clusters, which is about 5~pc. Further details on the timestep selection are discussed in the appendix.

@@

    Overall, across 50 simulations, we observe 73 gaps. The distribution of simulations with a given number of gaps is as follows: 
    \begin{itemize}
      \item 0 gaps in 4 simulations,
      \item 1 gap in 25 simulations,
      \item 2 gaps in 16 simulations,
      \item 3 gaps in 4 simulations,
      \item 4 gaps in 1 simulation. 
    \end{itemize} Excluding NGC~2808, we are left with 29 gaps, leading to an expected range of 0.6-1.5 gaps per simulation.
